Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-06-08 analysis/paper/paper.Rmd
INITIAL EDITS PER PLB COMMENTS

---
title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
author:
  - Josh M. London:
      email: josh.london@noaa.gov
      institute: [MML]
      correspondence: true
  - Paul B. Conn:
      institute: [MML]
      correspondence: false
  - Stacie K. Hardy:
      institute: [MML]
      correspondence: false
  - Erin L. Richmond:
      institute: [MML]
      correspondence: false
  - Jay M. Ver Hoef:
      institute: [MML]
      correspondence: false
  - Irina S. Trukhanova:
      institute: [MML]
      correspondence: false
  - Michael F. Cameron:
      institute: [MML]
      correspondence: false
  - Peter L. Boveng:
      institute: [MML]
      correspondence: false
institute:
  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
output: 
  officedown::rdocx_document:
    reference_docx: ../templates/jml_template.docx
    pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
bibliography: references.json
link-citations: true
csl: ../templates/royal-society-open-science.csl
always_allow_html: true
params:
   run_models: FALSE
---

@@

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring and
early summer (April-June) as sea ice begins to melt and retreat northward. Rapid
acceleration of climate change in Arctic ecosystems is therefore of concern as the
quantity and quality of suitable habitat is forecast to decrease. In this paper, we use
data collected from satellite-linked bio-loggers deployed between 2005 and 2020 to
investigate the seasonal timing and environmental factors affecting sea ice use by seals
(specifically, bearded, ribbon, and spotted seals) in the Bering and Chukchi seas. In
addition to providing baseline data on phenology, these data also allow us to refine
availability estimates needed to accurately estimate abundance from aerial survey counts
of seals basking on ice (i.e., to correct for the proportion of animals that are in the
water while surveys are conducted). Using generalized linear mixed pseudo-models to
properly account for temporal autocorrelation, we fit models with a variety of covariates
(e.g., day-of-year, solar hour, age-sex class, wind speed, barometric pressure,
temperature, precipitation) to examine their ability to explain variation in hourly
haul-out records. We found evidence for strong diurnal and within season patterns in
haul-out behavior, as well as strong weather effects (particularly wind and temperature).
In general, seals were more likely to rest on ice in the middle of the day and when wind
speed was low and temperatures were higher. Haul-out propensity increased through March
and April, peaking in May and early June before declining again. The timing and frequency
of haul-out events also varied based on species and age-sex class. For ribbon and spotted
seals, models with year effects were highly supported, indicating that the timing and
magnitude of haul-out behavior varied among years. However, we did not find evidence that
haul-out timing was linked annual sea ice extent. Our analysis emphasizes the importance
of accounting for variation in haul-out behavior when interpreting the number of seals
counted in aerial surveys.

@@

The global climate crisis is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic ecosystems, taxa, and the human communities who live
in the region [@huntington2020ncc]. Such disruptions are a particular cause of concern for
ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus barbatus*), spotted
(*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that depend on spring and
early summer sea ice (March-June) in the Bering and Chukchi seas as a platform for
important functions, such as pupping, nursing, breeding, and molting [@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps prevent our
ability to make definitive predictions about the ultimate effects of changes in sea ice on
the behavior, abundance, and distribution of these seals. Knowledge about biological
constraints on the phenology of reproductive and molting behavior is generally lacking, so
it is difficult to predict the readiness with which ice-associated seal species will be
able to adapt to future changes (e.g., by adjusting pupping or molting schedules to
earlier dates or different locales). Additionally, trends in abundance of these species
are largely unknown, so it is difficult to say what effect, if any, declines in sea ice
habitat have had, or will have, on seal densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remote nature of the Bering and
Chukchi seas mean direct observation of seal behavior is limited. Bio-logging devices are,
thus, especially useful tools for collecting key information on movement and haul-out
behavior for these species. Records of time spent out of the water provide valuable data
for identifying covariates that explain variation in haul-out behavior. For instance, in
the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied on bio-logging data to
demonstrate greater haul-out propensity in juvenile crabeater seals (*Lobodon
carcinophaga*) than adults, with highest probabilities in February and at times close to
solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson et al.
[-@bengtson2005pb] documented a higher propensity for ringed seal basking near solar noon,
as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon , and spotted
seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that the
proportion of time bearded seals spent hauled out progressively increased through spring
and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities increased
gradually starting in March and peaking in May and June for bearded, ribbon, and spotted
seals. However, their primary focus was to develop a correction factor for estimating seal
abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee], they did not
attempt to separately analyze haul-out schedules for different age and sex classes
(although initial models with sex and age as fixed effects were insignificant at p = 0.05
[@verhoef2014sm]).

>>>>

---
title: "Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas"
author:
  - Josh M. London:
      email: josh.london@noaa.gov
      institute: [MML]
      correspondence: true
  - Paul B. Conn:
      institute: [MML]
      correspondence: false
  - Stacie K. Hardy:
      institute: [MML]
      correspondence: false
  - Erin L. Richmond:
      institute: [MML]
      correspondence: false
  - Jay M. Ver Hoef:
      institute: [MML]
      correspondence: false
  - Michael F. Cameron:
      institute: [MML]
      correspondence: false
  - Peter L. Boveng:
      institute: [MML]
      correspondence: false
institute:
  - MML: Marine Mammal Laboratory, Alaska Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA
output: 
  officedown::rdocx_document:
    reference_docx: ../templates/jml_template.docx
    pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
bibliography: references.json
link-citations: true
csl: ../templates/royal-society-open-science.csl
always_allow_html: true
params:
   run_models: FALSE
---

@@

Ice-associated seals rely on sea ice for a variety of activities, including breeding,
molting, pupping, and resting. In the Arctic, many of these activities occur in spring
(April-June) as sea ice begins to melt and retreat northward. Rapid acceleration of
climate change in Arctic ecosystems is therefore of concern as the quantity and quality of
suitable habitat is forecast to decrease. In this paper, we use data collected from
satellite-linked bio-loggers deployed between 2005 and 2020 to investigate the seasonal
timing and environmental factors affecting sea ice use by seals (specifically, bearded,
ribbon, and spotted seals) in the Bering and Chukchi seas. In addition to providing
baseline data on phenology, these data also allow us to quantify ‘availability’, which is
needed to accurately estimate abundance from aerial survey counts of seals basking on ice
(i.e., to correct for the proportion of animals that are in the water while surveys are
conducted). Using generalized linear mixed pseudo-models to properly account for temporal
autocorrelation, we fit models with a variety of covariates (e.g., day-of-year, solar
hour, age-sex class, wind speed, barometric pressure, temperature, precipitation) to
examine their ability to explain variation in hourly haul-out records. We found evidence
for strong diurnal and within-season patterns in haul-out behavior, as well as strong
weather effects (particularly wind and temperature). In general, seals were more likely to
rest on ice in the middle of the day and when wind speed was low and temperatures were
higher. Haul-out propensity increased through March and April, peaking in May and early
June before declining again. The timing and frequency of haul-out events also varied based
on species and age-sex class. For ribbon and spotted seals, models with year effects were
highly supported, indicating that the timing and magnitude of haul-out behavior varied
among years. However, we did not find evidence that haul-out timing was linked to annual
sea ice extent. Our analysis emphasizes the importance of accounting for variation in
haul-out behavior when interpreting the number of seals counted in aerial surveys.

@@

The global climate disruption is causing considerable reduction in Arctic sea ice extent,
volume, and seasonal presence [@kwok2018erl, @meier2014rg, @wang2017dsrpitsio]. These
changes have ripple effects on Arctic organisms, ecosystems, and the human communities who
live in the region [@huntington2020ncc]. Such disruptions are a particular cause of
concern for ice-associated seals (ringed (*Phoca hispida*), bearded (*Erignathus
barbatus*), spotted (*Phoca largha*), and ribbon seals (*Histriophoca fasciata*)) that
depend on spring and early summer sea ice (March-June) in the Bering and Chukchi seas as a
platform for important functions, such as pupping, nursing, breeding, and molting
[@boveng2009a;
@boveng2013; @cameron2010a; @kelly2010]. Limited data and large knowledge gaps complicate
predictions about the ultimate effects of changes in sea ice on the behavior, abundance,
and distribution of these seals. Knowledge about evolutionary constraints on the timing of
reproductive and molting behavior is generally lacking, so it is difficult to predict the
readiness with which ice-associated seal species will be able to adapt to future changes
(e.g., by adjusting pupping or molting schedules to earlier dates or different locales).
Additionally, trends in abundance of these species are largely unknown, so it is difficult
to say what effect, if any, declines in sea ice habitat have had, or will have, on seal
densities.

For ice-associated seals, haul out is tightly linked with sea ice. Ribbon seals rest out
of the water almost exclusively on sea ice and are mostly pelagic outside the spring
pupping, breeding, and molting season [@burns1981]. While spotted and bearded seals rest
on beaches during ice-free months, they strongly prefer sea ice as a resting platform
during the spring and early summer [@burns2009]. The remoteness of the Bering and Chukchi
seas mean direct scientific observation of seal behavior is impractical. Bio-logging
devices are, thus, especially useful tools for collecting key information on movement and
haul-out behavior for these species. Bio-logging records of time spent out of the water
provide valuable data for identifying covariates that explain variation in haul-out
behavior. For instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004pb] relied
on bio-logging data to demonstrate greater haul-out propensity in juvenile crabeater seals
(*Lobodon carcinophaga*) than adults, with highest probabilities in February and at times
close to solar noon. In the Arctic, Von Duyke et al. [-@vonduyke2020ee] demonstrated
seasonal changes between diurnal and nocturnal haul-out behavior of ringed seals. Bengtson
et al. [-@bengtson2005pb] documented a higher propensity for ringed seal basking near
solar noon, as did Ver Hoef et al. [-@verhoef2014sm] in an analysis of bearded , ribbon ,
and spotted seals using much larger sample sizes. Olnes et al. [-@olnes2020pb] showed that
the proportion of time bearded seals spent hauled out progressively increased through
spring and summer. And, Ver Hoef et al. [-@verhoef2014sm] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded, ribbon, and
spotted seals. However, their primary focus was to develop a correction factor for
estimating seal abundance, and like a subsequent analysis by Conn et al. [-@conn2014mee],
they did not attempt to separately analyze haul-out schedules for different age and sex
classes (although initial models with sex and age as fixed effects were insignificant at p
= 0.05 [@verhoef2014sm]).
